The Proofs

### 2023-11-13 proofs/tex/norm-lemmas.tex
Restructure proofs

\begin{lemma}
  \label{lemma:equiv-fv}
  Set of free variables is invariant under equivalence.
  \begin{itemize}
  \item[$-$] If $[[iN ≈ iM]]$ then $[[fv iN]] \equiv [[fv iM]]$ (as sets)
  \item[$+$] If $[[iP ≈ iQ]]$ then $[[fv iP]] \equiv [[fv iQ]]$ (as sets)
  \end{itemize}
\end{lemma}
\begin{proof}
  Straightforward mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$
\end{proof}

\begin{lemma}
  \label{lemma:fv-nf}
  Free variables are not changed by the normalization
  \begin{itemize}
  \item[$-$] $[[fv iN]] \equiv [[fv nf(iN)]]$
  \item[$+$] $[[fv iP]] \equiv [[fv nf(iP)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  By straightforward induction on $[[iN]]$ and mutually on $[[iP]]$.
\end{proof}

\begin{lemma}[Soundness of normalization]
  \label{lemma:normalization-soundness}
  \hfill
  \begin{itemize}
    \item[$-$] $[[iN ≈ nf(iN)]]$
    \item[$+$] $[[iP ≈ nf(iP)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[nf(iN) = iM]]$ and $[[nf(iP) = iQ]]$.
  Let us consider how this judgment is formed:
  \begin{caseof}
    \item{\nameref{\ottdruleNrmNVarLabel} and \nameref{\ottdruleNrmPVarLabel}}\\ By
      the corresponding equivalence rules.
    \item{\nameref{\ottdruleNrmShiftULabel}, \nameref{\ottdruleNrmShiftDLabel},
        and \nameref{\ottdruleNrmArrowLabel}}\\
      By the induction hypothesis and the corresponding congruent equivalence rules.
    \item{\nameref{\ottdruleNrmForallLabel}}, i.e. $[[nf(∀pas.uN) = ∀pas'.uN']]$ \label{case:norm-soundness:forall}\\
      From the induction hypothesis, we
      know that $[[iN ≈ iN']]$. In particular, by \cref{lemma:equiv-fv}, $[[fv
        iN]] \equiv [[fv iN']]$.
      Then by \cref{lemma:ord-soundness}, $[[{pas'}]]
      \equiv [[{pas} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$, and thus,
      $[[{pas'} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$.

@@

\begin{corollary}[Normalization preserves ordering]
  \label{corollary:normalization-ord}
  For any $[[varset]]$,
  \begin{itemize}
  \item[$-$] $[[ord varset in nf(iN)]] = [[ord varset in iM]]$
  \item[$+$] $[[ord varset in nf(iP)]] = [[ord varset in iQ]]$
  \end{itemize}
\end{corollary}
\begin{proof}
  Immediately from \cref{lemma:ord-completeness,lemma:normalization-soundness}.
\end{proof}

\begin{lemma}[Distributivity of normalization over substitution]
  \label{lemma:norm-subst-distr} Normalization of a term distributes over substitution.
  Suppose that $[[σ]]$ is a substitution, $[[iN]]$ and $[[iP]]$ are types. Then
    \begin{itemize}
      \item[$-$] $[[nf([σ]iN)]] = [[ [nf(σ)] nf(iN) ]]$
      \item[$+$] $[[nf([σ]iP)]] = [[ [nf(σ)] nf(iP) ]]$
  \end{itemize}
  where $[[nf(σ)]]$ means pointwise normalization: $[[ [nf(σ)] α⁻]] = [[nf([σ]α⁻)]]$.
\end{lemma}
\begin{proof}
  Mutual induction on $[[iN]]$ and $[[iP]]$.
  \begin{caseof}
    \item $[[iN]]$ = $[[na]]$ \\
      \label{case:norm-subst-distr-var}
      $[[nf([σ]iN)]] = [[ nf([σ]na) ]] = [[ [nf(σ)]na ]] $.

@@

     To show alpha-equivalence of 
     $[[ [nf(σ)] ∀pas'.nf(iM) ]]$ and $[[ ∀pas'.[nf(σ)]nf(iM) ]]$,
     we can assume that $[[{pas'} ∩ Γ1]] = \emptyset$, and $[[{pas'} ∩ Γ2]]
     = \emptyset$.

@@

\begin{corollary}[Commutativity of normalization and renaming]
  \label{lemma:norm-subst-commute} Normalization of a term commutes with renaming.
  Suppose that $\mu$ is a bijection between two sets of variables
  $\mu : A \leftrightarrow B$. Then
  \begin{itemize}
    \item[$-$] $[[nf([mu]iN)]] = [[ [mu] nf(iN) ]]$
    \item[$+$] $[[nf([mu]iP)]] = [[ [mu] nf(iP) ]]$
  \end{itemize}
\end{corollary}
\begin{proof}
  Immediately from \cref{lemma:norm-subst-distr},
  after noticing that $[[nf(mu)]] = [[mu]]$.
\end{proof}

\begin{lemma}[Completeness of quantified normalization]
  \label{lemma:normalization-completeness}
  Normalization returns the same representative for equivalent types.

  \begin{itemize}
  \item[$-$] If $[[iN ≈ iM]]$ then $[[nf(iN)]] = [[nf(iM)]]$
  \item[$+$] If $[[iP ≈ iQ]]$ then $[[nf(iP)]] = [[nf(iQ)]]$
  \end{itemize}
\end{lemma}

\begin{proof}
  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$.
  \begin{caseof}
  \item {\nameref{\ottdruleEOneForallLabel}} \label{case:ord-completeness:forall} \\

@@

\begin{lemma}[Idempotence of normalization]
  \label{lemma:norm-idemp}
  Normalization is idempotent
  \begin{itemize}
  \item[$-$] $[[nf(nf(iN))]] = [[nf(iN)]]$
  \item[$+$] $[[nf(nf(iP))]] = [[nf(iP)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  By applying \cref{lemma:normalization-completeness} to \cref{lemma:normalization-soundness}.
\end{proof}

\begin{lemma}
  \label{lemma:normal-after-subst}
  The result of a substitution is normalized if and only if the initial type and
  the substitution are normalized.

  Suppose that $[[σ]]$ is a substitution  $[[Γ2 ⊢ σ : Γ1]]$,
  $[[iP]]$ is a positive type ($[[Γ1 ⊢ iP]]$),
  $[[iN]]$ is a negative type ($[[Γ1 ⊢ iN]]$). Then
  \begin{itemize}
  \item[$+$]
    $[[ [σ]iP  ]] \text{ is normal} \iff
    \begin{cases}
      [[σ|fv(iP)]] &\text{is normal}\\
      [[iP]]       &\text{is normal}\\
    \end{cases} $
  \item[$-$]
    $[[ [σ]iN  ]] \text{is normal} \iff
    \begin{cases}
      [[σ|fv(iN)]] &\text{is normal}\\
      [[iN]]       &\text{is normal}\\
    \end{cases} $
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[Γ1 ⊢ iP]]$ and $[[Γ1 ⊢ iN]]$.
  \begin{caseof}
  \item $[[iN]] = [[na]]$\\
    Then $[[iN]]$ is always normal, and
    the normality of $[[σ|{na}]]$ by the definition means $[[ [σ]na ]]$ is normal.

@@

>>>>

\obsNormDeterministic*
\begin{proof}
  By straightforward induction using \cref{obs:ord-deterministic}.
\end{proof}

\lemmaFvNf*
\begin{proof}
  By mutual induction on $[[iN]]$ and $[[iP]]$.
  The base cases 
  (\ruleref{\ottdruleNrmNVarLabel} and \ruleref{\ottdruleNrmPVarLabel})
  are trivial; the congruent cases
  (\ruleref{\ottdruleNrmShiftULabel},
  \ruleref{\ottdruleNrmShiftDLabel}, and
  \ruleref{\ottdruleNrmArrowLabel}) are proved by the induction hypothesis.

  Let us consider the case when the term is formed by $[[∀]]$,
  that is the normalization judgment has a shape 
  $[[nf(∀pas.iN) = ∀pas'.iN']]$,
  where by inversion $[[nf(iN) = iN']]$
  and $[[ord {pas} in iN' = pas']]$.
  By the induction hypothesis, $[[fv iN = fv iN']]$.
  Since $[[fv(∀pas.iN) = fv iN \ {pas}]]$,
  and $[[fv(∀pas'.iN') = fv iN' \ {pas'}]]$,
  it is left to show that $[[fv iN \ {pas} = fv iN \ {pas'}]]$.
  By \cref{lemma:ord-completeness}, 
  $[[{pas'}]] = [[{pas} ∩ fv iN']] = [[{pas} ∩ fv iN]]$.
  Then
  $[[fv iN \ {pas} = fv iN \ ({pas} ∪ fv iN)]]$ by
  set-theoretic properties, and
  thus, $[[fv iN \ {pas} = fv iN \ {pas'}]]$.

  The case when the term is positive and formed by $[[∃]]$ is symmetric.
\end{proof}

\lemmaNormSoundness*
\begin{proof}
  Mutual induction on $[[nf(iN) = iM]]$ and $[[nf(iP) = iQ]]$.
  Let us consider how this judgment is formed:
  \begin{caseof}
    \item{\nameref{\ottdruleNrmNVarLabel} and \nameref{\ottdruleNrmPVarLabel}}\\ By
      the corresponding equivalence rules.
    \item{\nameref{\ottdruleNrmShiftULabel}, \nameref{\ottdruleNrmShiftDLabel},
        and \nameref{\ottdruleNrmArrowLabel}}\\
      By the induction hypothesis and the corresponding congruent equivalence rules.
    \item{\nameref{\ottdruleNrmForallLabel}}, i.e. $[[nf(∀pas.uN) = ∀pas'.uN']]$ \label{case:norm-soundness:forall}\\
      From the induction hypothesis, we
      know that $[[iN ≈ iN']]$. In particular, by \cref{lemma:equiv-fv}, $[[fv
        iN]] \equiv [[fv iN']]$.
      Then by \cref{lemma:ord-soundness}, $[[{pas'}]]
      \equiv [[{pas} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$, and thus,
      $[[{pas'} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$.

@@

\corollaryWfNf*
\begin{proof}
  Immediately from \cref{lemma:wf-equiv,lemma:normalization-soundness}.
\end{proof}

\corollaryWfSNf*
\begin{proof}
  Let us prove the forward direction.
  Suppose that $[[α± ∊ Γ1]]$.  Let us show that $[[Γ2 ⊢ [nf(σ)]α±]]$.
  By the definition of substitution normalization,
  $[[ [nf(σ)]α± = nf([σ]α±) ]]$. Then by \cref{corollary:wf-nf},
  to show $[[Γ2 ⊢ nf([σ]α±)]]$, it suffices to show $[[Γ2 ⊢ [σ]α±]]$,
  which holds by the assumption $[[Γ2 ⊢ σ : Γ1]]$.

  The backward direction is proved analogously.
\end{proof}

\lemmaNormSubstSig*
\begin{proof}
  Suppose that $[[α± ∊ Γ1]]$. 
  Then by \cref{corollary:wf-nf}, $[[Γ2 ⊢ nf([σ]α±)]] = [[ [nf(σ)]α± ]]$ 
  is equivalent to $[[Γ2 ⊢ [σ]α±]]$.

  Suppose that $[[α± ∉ Γ1]]$. 
  $[[Γ2 ⊢ σ : Γ1]]$ means that $[[ [σ]α± = α± ]]$, 
  and then $[[ [nf(σ)]α± ]] = [[nf([σ]α±)]] = [[nf(α±)]] = [[α±]]$.
\end{proof}

\corollaryNfSoundWrtSubtEquiv*
\begin{proof}
  Immediately from \cref{lemma:normalization-soundness,corollary:wf-nf,lemma:equiv-soundness}.  
\end{proof}

\corollaryNfPresSubt*
\begin{proof}
  \hfill
  \begin{itemize}
    \item [$+$]  
    \begin{itemize}
      \item [$\Rightarrow$] Let us assume $[[Γ ⊢ iP ≥ iQ]]$.
        By \cref{corollary:nf-sound-wrt-subt-equiv},
        $[[Γ ⊢ iP ≈ nf(iP)]]$ and $[[Γ ⊢ iQ ≈ nf(iQ)]]$, 
        in particular, by inversion, 
        $[[Γ ⊢ nf(iP) ≥ iP]]$ and $[[Γ ⊢ iQ ≥ nf(iQ)]]$.
        Then by transitivity of subtyping 
        (\cref{lemma:subtyping-transitivity}), 
        $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
      \item [$\Leftarrow$] Let us assume $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
        Also by \cref{corollary:nf-sound-wrt-subt-equiv}
        and inversion, 
        $[[Γ ⊢ iP ≥ nf(iP)]]$ and $[[Γ ⊢ nf(iQ) ≥ iQ]]$.
        Then by the transitivity, $[[Γ ⊢ iP ≥ iQ]]$.
    \end{itemize}
    \item [$-$] The negative case is proved symmetrically.
  \end{itemize}
\end{proof}

\corollaryNormPreservesOrdering*
\begin{proof}
  Immediately from \cref{lemma:ord-completeness,lemma:normalization-soundness}.
\end{proof}

\lemmaNormSubstDistr*
\begin{proof}
  Mutual induction on $[[iN]]$ and $[[iP]]$.
  \begin{caseof}
    \item $[[iN]]$ = $[[na]]$ \\
      \label{case:norm-subst-distr-var}
      $[[nf([σ]iN)]] = [[ nf([σ]na) ]] = [[ [nf(σ)]na ]] $.

@@

     To show the alpha-equivalence of 
     $[[ [nf(σ)] ∀pas'.nf(iM) ]]$ and $[[ ∀pas'.[nf(σ)]nf(iM) ]]$,
     we can assume that $[[{pas'} ∩ Γ1]] = \emptyset$, and $[[{pas'} ∩ Γ2]]
     = \emptyset$.

@@

\lemmaNormSubstCommute*
\begin{proof}
  Immediately from \cref{lemma:norm-subst-distr}, after noticing that $[[nf(mu)]] = [[mu]]$.
\end{proof}

\lemmaNormalizationCompleteness*
\begin{proof}
  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$.
  \begin{caseof}
  \item {\nameref{\ottdruleEOneForallLabel}} \label{case:ord-completeness:forall} \\

@@

\lemmaDeclEquivAlg*
\begin{proof} \hfill
  \begin{itemize}
    \item[$+$] Let us prove both directions separately.
    \begin{itemize}
      \item[$\Rightarrow$] 
        exactly by \cref{lemma:normalization-completeness},
      \item[$\Leftarrow$] 
        from \cref{lemma:normalization-soundness}, we know
        $[[iP ≈ nf(iP)]] = [[nf(iQ) ≈ iQ]]$, then by transitivity (\cref{lemma:decl-equiv-transitivity}),
        $[[iP ≈ iQ]]$.
    \end{itemize}
    \item[$-$] For the negative case, the proof is the same.
  \end{itemize}
\end{proof}

\corollaryNfCompleteWrtSubtEquiv*
\begin{proof}
  Immediately from \cref{lemma:equiv-completeness,lemma:normalization-completeness}.
\end{proof}

\lemmaNormIdemp*
\begin{proof}
  By applying \cref{lemma:normalization-completeness} to \cref{lemma:normalization-soundness}.
\end{proof}

\lemmaNormalAfterSubst*
\begin{proof}
  Mutual induction on $[[Γ1 ⊢ iP]]$ and $[[Γ1 ⊢ iN]]$.
  \begin{caseof}
  \item $[[iN]] = [[na]]$\\
    Then $[[iN]]$ is always normal, and
    the normality of $[[σ|{na}]]$ by the definition means $[[ [σ]na ]]$ is normal.

@@

\lemmaSubtInducedEquivAlg*
\begin{proof}
  Let us prove the positive case, the negative case is symmetric.
  We prove both directions of $\iff$ separately:
  \begin{itemize}
    \item [$\Rightarrow$] exactly \cref{corollary:nf-complete-wrt-subt-equiv};
    \item [$\Leftarrow$] by \cref{lemma:decl-equiv-algorithmization,lemma:equiv-soundness}.
  \end{itemize}
\end{proof}

\corSubstPresDeclEquiv*
\begin{proof}
  $
  \begin{aligned}[t] 
    [[ iP ≈ iQ ]] &\Rightarrow        [[ nf(iP) = nf(iQ) ]]
                  && \text{by \cref{lemma:subt-equiv-algorithmization}}\\
                  &\Rightarrow [[ [nf(σ)]nf(iP) = [nf(σ)]nf(iQ)]]\\
                  &\Rightarrow [[ nf([σ]iP) = nf([σ]iQ)]]
                  && \text{by \cref{lemma:norm-subst-distr}}\\ 
                  &\Rightarrow        [[ [σ]iP ≈ [σ]iQ ]]
                  && \text{by \cref{lemma:subt-equiv-algorithmization}}\\
  \end{aligned} 
  $
\end{proof}
